\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-07-08 scratch/notes_john/JSS-article/JSS-article-4.Rmd
added JSS-article-4.Rmd and .pdf under scratch\notes_john. Edited Secs. 1-3.

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
palette(car::carPalette())
library(car)     # included this to avoid scatterplotMatrix() error
```

@@

Validating research by replication on independently collected data is a common scientific norm. Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data. Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples. See @Harrell:2015 [Sec. 5.3] for this and other remarks about data-division and cross-validation.

@@

* The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.

@@

```{r mpg-horsepower-scatterplot-polynomials, echo=FALSE}
#| out.width = "50%",
#| fig.height = 5,
#| fig.cap = "\\code{mpg} vs. \\code{horsepower} for the \\code{Auto} data, showing fitted polynomials of degree 1 through 5."
plot(mpg ~ horsepower, data=Auto)
horsepower <- with(Auto, 
                   seq(min(horsepower), max(horsepower), 
                       length=1000))
for (p in 1:5){
  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
  lines(horsepower, mpg, col=p + 1, lty=p, lwd=4)
}
legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=4,
       title="Degree", inset=0.02)
```

We'll focus here on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}. The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}. The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).

Figure \ref{fig:mpg-horsepower-MSE-se} shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$. The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.

```{r mpg-horsepower-MSE-se, echo=FALSE}
#| out.width = "50%",
#| fig.height = 5,
#| fig.cap = "Estimated squared error as a function of polynomial degree, $p$"

@@

@@

* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).

* \code{ncores}, the number of cores to employ for parallel computation; if \code{cores = 1} (the default), the computations are not parallelized.

To illustrate, we perform 10-fold CV for a quadratic polynomial fit to the \code{Auto} data:
```{r cv-lm-1}
m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
summary(m.auto)

@@

```{r cv.lm-2`}
cv(m.auto, k="loo")
```
The \code{"hatvalues"} method reports only the CV estimate of MSE. Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted. All three methods produce exact results for a linear model (within the precision of floating-point computations):
```{r cv.lm-3}
cv(m.auto, k="loo", method="naive", confint=TRUE)

cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
```
The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE (and a confidence interval around it) along with the full-sample MSE, but bias isn't an issue for LOO CV.

@@

```{r, polynomial-regression-CV-graph-2, fig.show="hold"}
#| fig.height = 5,
#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$.",
#| fig.subcap=c('', ''),
#| fig.ncol = 2, out.width = "50%", fig.align = "center"
plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
```

In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-MSE-se} on page \pageref{fig:mpg-horsepower-MSE-se}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.

@@

[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The \code{lmer()} function in the \pkg{lme4} package, for example, is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects. We present an example of the latter in a vignette for the \pkg{cv} package.

@@

Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
```{r HSB-lmer, cache=TRUE}
library("lme4", quietly=TRUE)
hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
                   + (cses | school), data=HSB)
summary(hsb.lmer, correlation=FALSE)
```

@@

The four data sets differ in the relative magnitude of between-cluster variance compared with within-cluster variance. They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.

For these analyses, we use the `glmmTMB()` function in the **glmmTMB** package [@BrooksEtAl] because, in our experience, it is more likely to converge than functions in the **nlme** and the **lme4** packages for models with low between-cluster variance.

Consider a researcher studying the effect of the dosage of a drug on the severity of symptoms for a hypothetical disease. The researcher has longitudinal data on 20 patients, each of whom was observed on five occasions in which patients received different dosages of the drug.  The data are observational, with dosages prescribed by the patients' physicians, so that patients who were more severely affected by the disease received higher dosages of the drug.

@@

To construct the four data sets, we first construct a template data set from a population with a between patient effect of dosage of $1.0$ and a within-patient effect of $-0.5$. We then apply four different multipliers to the between-patient variation from the regression line of patient centroids. The resulting data sets, shown in Figure \ref{fig:plot1}, differ in the between-patient variance of patient centroids from a common between-patient regression line, exhibiting a progression from low to high variance around the common regression line.[^lattice]

Note that each patient has the same relative configuration of dosages and symptoms in each of the four data sets. To help visualize the structure of the data, estimated 50% concentration ellipses are shown for each patient.

Using such data, the researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models are sequentially nested with fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`. The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`. We thus fit four models to four data sets for a total of 16 models.

```{r include=FALSE, echo=FALSE}
library("glmmTMB") # necessary for some reason to knit vignette in RStudio, harmless otherwise
```
```{r parameters, include=FALSE}
# Parameters:
Nb <- 20     # number of patients
Nw <- 5      # number of occasions for each patient
Bb <- 1.0    # between-patient regression coefficient on patient means
Bw <- -0.5   # within-patient effect of x

@@

[^lattice]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.

```{r plot1, echo=FALSE}
#| out.width = "100%",
#| fig.height = 3.0,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD. The ellipses are estimated 50% concentration ellipses for each patient. The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
library("lattice")
library("latticeExtra")
layer <- latticeExtra::layer # to avert conflicts with ggplot2::layer
xyplot(y ~ x | SD_ratio, data = Data, group = patient,
       ylab = list('symptoms', cex = 0.7),
       xlab = list('dosage', cex = 0.7),
       par.strip.text = list(cex = 0.7),
       scales = list(cex = 0.7, x = list(alternating = FALSE)),
       layout = c(Nv, 1),
       par.settings = list(superpose.symbol = list(pch = 1, cex = 0.7))) +
  layer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                      level = 0.5),
        panel.abline(a = 10, b = 1))
```

```{r model-fits,include=FALSE}
model.formulas <- c(
  ' ~ 1'             =  y ~ 1 + (1 | patient),
  '~ 1 + x'          =  y ~ 1 + x + (1 | patient),
  '~ 1 + x + xm'     =  y ~ 1 + x + xm + (1 | patient),
  '~ 1 + I(x - xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
)
fits <- lapply(split(Data, ~ SD_ratio),
               function(d) {
                 lapply(model.formulas, function(form) {
                   glmmTMB(form, data = d)
                 })
               })
```
We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or "BLUPs"---as would be used for cross-validation based on cases (i.e., occasions within patients):
```{r predict,include=FALSE}
pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  
pred.BLUPs <- lapply(fits, lapply, predict)
```
```{r data-predictions,include=FALSE}
Dataf <- lapply(split(Data, ~ SD_ratio),
                function(d) {
                  lapply(names(model.formulas), 
                         function(form) cbind(d, formula = form))
                }) |> 
  lapply(function(dlist) do.call(rbind, dlist)) |> 
  do.call(rbind, args = _) |> 
  within(
    {
      pred.fixed <- unlist(pred.fixed)
      pred.BLUPs <- unlist(pred.BLUPs)
      panel <- factor(formula, levels = c(names(model.formulas), 'data'))
    }
  )
Datap <- reshape(Dataf, direction = "long", sep = '.',
                 varying = c('pred.fixed', 'pred.BLUPs'),
                 timevar = 'prediction_type')
Datap$panel2 <- with(Datap, paste0(formula,":",prediction_type))
flevels <- c(unique(Datap$panel2), 'data')
Datap$panel2 <- factor(Datap$panel2, levels = flevels)
Data$panel <- factor('data', levels = c(names(model.formulas), 'data'))
Data$panel2 <- factor('data', levels = flevels)
```
The fixed- and random-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-combined} along with a summary of the data using 50% concentration ellipses.

@@

{   xyplot(y ~ x | SD_ratio * panel3, Data,
         groups = patient, type = 'n',
         drop.unused.levels = F,
         par.strip.text = list(cex = 0.4),
         scales = list(x = list(alternating = FALSE, cex = 0.7),
                       y = list(alternating =2, cex = 0.5)),
         xlab = list('dosage', cex = 0.7),
         between =
             list(y = c(0,0,0,0.25,0,0,0,.25)),
         ylab = list(
           paste0(
             paste0(rep(' ',18),  collapse = ''),
                  'fixed-effects',
             paste0(rep(' ',52), collapse = ''),
             'BLUPs',
             paste0(rep(' ',32), collapse = ''),
             'data'),
           cex = 0.7)) +
    glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,
                         level = 0.5),
           panel.abline(a = 10, b = 1)) +
    xyplot(pred  ~ x | SD_ratio * panel2, Datap, type = 'l',
           groups = patient,
           drop.unused.levels = F)
} -> re.plot
useOuterStrips(re.plot)

@@

Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`. The inclusion of a contextual variable in the model corrects this problem.

Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform with the within-patient slopes for the `~ 1 + x` model in the two data sets with the smallest between-patient variances.

For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variance.  With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.

@@

The third model, `~ 1 + x + xm`, includes a contextual effect of `x`---that is, the cluster mean `xm`---along with `x` and the intercept in the fixed-effect part of the model, along with a random intercept. This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | patient)`, which is the model that generated the data.  The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects only model with `x` and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).

We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only. In order to reduce between-model random variability in comparisons of models on the same data set, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.

>>>>

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
options(digits=5)
palette(car::carPalette())
library(car)     # included this to avoid scatterplotMatrix() error
```

@@

Validating research by replication on independently collected data is a common scientific norm. Emulating this process in a single study by data-division is less common [see, e.g., @Barnard:1974]: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data. Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples. See @Harrell:2015 [Sec. 5.3] for this and other remarks about data-division and cross-validation.

@@

* The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a potentially complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.

@@

```{r mpg-horsepower-polynomials, fig.show="hold", echo=FALSE}
#| out.width = "45%",
#| fig.height = 5,
#| fig.cap = "(a) \\code{mpg} vs. \\code{horsepower} for the \\code{Auto} data, showing fitted polynomials of degree 1 through 5. (b) Estimated squared error as a function of polynomial degree, $p = 1, \\ldots, 10$.",
#| fig.subcap=rep("", 2)
plot(mpg ~ horsepower, data=Auto)
horsepower <- with(Auto, 
                   seq(min(horsepower), max(horsepower), 
                       length=1000))
for (p in 1:5){
  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
  lines(horsepower, mpg, col=p + 1, lty=p, lwd=4)
}
legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=4,
       title="Degree", inset=0.02)

@@

We'll focus on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-polynomials} (a). The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-polynomials} (a). The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).

Figure \ref{fig:mpg-horsepower-polynomials} (b) shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$. The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.

@@

* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} (the default) selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).

* \code{ncores}, the number of cores to employ for parallel computation; if \code{ncores = 1} (the default), the computations are not parallelized.

To illustrate, we perform 10-fold CV for a quadratic polynomial fit to the \code{Auto} data:[^car-package]

[^car-package]: We load the \pkg{car} package [@FoxWeisberg:2019] here for the \code{brief()} function and to use below.

```{r cv-lm-1}
library("car", quietly=TRUE)

m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
brief(m.auto)

@@

```{r cv.lm-2`}
cv(m.auto, k="loo")
```
The \code{"hatvalues"} method reports only the CV estimate of MSE. Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted. All three methods produce exact results for a linear model (within the precision of floating-point computations); for example:
```{r cv.lm-3}
cv(m.auto, k="loo", method="naive", confint=TRUE)
```
The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE (and a confidence interval around it) along with the full-sample MSE, but bias isn't an issue for LOO CV.

@@

```{r, polynomial-regression-CV-graph-2, fig.show="hold"}
#| fig.height = 5,
#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p = 1, 2, \\ldots, 10$.",
#| fig.subcap=c('', ''),
#| fig.ncol = 2, out.width = "50%", fig.align = "center"
plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
```

In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-polynomials} (b) on page \pageref{fig:mpg-horsepower-polynomials}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.

@@

[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The \code{lmer()} function in the \pkg{lme4} package, for example, is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects. We present an example of crossed random effects in a vignette for the \pkg{cv} package.

@@

Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data (summarizing the model with the \code{S()} function from the \pkg{car} package):
```{r HSB-lmer, cache=TRUE}
library("lme4", quietly=TRUE)
hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
                   + (cses | school), data=HSB)
S(hsb.lmer, brief=TRUE)
```

@@

The four data sets differ in the magnitude of between-cluster variance compared to within-cluster variance. They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.[^code-in-complements]

[^code-in-complements]: The purpose of this section is to explain some of the complexities of cross-validating mixed-effects models, which motivates the approach that we take for mixed models in the \pkg{cv} package. The examples in this section introduce no features of the \pkg{cv} package that we haven't already encountered, and the generation of data for these examples involves unenlightening data management. Consequently, we don't show the code for data generation, model fitting, and graphs in this section; the code is in the complementary materials for the paper, however. For these analyses, we used the `glmmTMB()` function in the \pkg{glmmTMB} package [@BrooksEtAl] because, in our experience, it is more likely to converge than functions in the \pkg{nlme} and the \pkg{lme4} packages for models with low between-cluster variance. We used the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for the graphs in this section.

Consider a researcher studying the effect of the dosage of a drug on the severity of symptoms for a hypothetical disease. The researcher has longitudinal data on 20 patients, each of whom was observed on 5 occasions in which patients received different dosages of the drug.  The data are observational, with dosages prescribed by the patients' physicians, so that patients who were more severely affected by the disease received higher dosages of the drug.

@@

To generate the four data sets, we first constructed a template data set from a population with a between-patient effect of dosage of $1.0$ and a within-patient effect of $-0.5$. We then applied four different multipliers to the between-patient variation from the regression line of patient centroids. The resulting data sets, shown in Figure \ref{fig:plot1}, differ in the between-patient variance of patient centroids from a common between-patient regression line, exhibiting a progression from low to high variance around the common regression line. Each patient has the same relative configuration of dosages and symptoms in each of the four data sets. To help visualize the structure of the data, estimated 50% Gaussian concentration ellipses are shown for each patient.

```{r include=FALSE, echo=FALSE}
library("glmmTMB") # necessary for some reason to knit vignette in RStudio, harmless otherwise
```
```{r parameters, include=FALSE, cache=TRUE}
# Parameters:
Nb <- 20     # number of patients
Nw <- 5      # number of occasions for each patient
Bb <- 1.0    # between-patient regression coefficient on patient means
Bw <- -0.5   # within-patient effect of x

@@

```{r plot1, echo=FALSE, cache=TRUE}
#| out.width = "75%",
#| fig.height = 2.5,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD. The ellipses are estimated 50% concentration ellipses for each patient. The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel, where $y$ is symptoms and $x$ is dosage."
library("lattice")
library("latticeExtra")
layer <- latticeExtra::layer # to avert conflicts with ggplot2::layer
xyplot(y ~ x | SD_ratio, data = Data, group = patient,
       ylab = list('symptoms (y)', cex = 0.7),
       xlab = list('dosage (x)', cex = 0.7),
       par.strip.text = list(cex = 0.7),
       scales = list(cex = 0.7, x = list(alternating = FALSE)),
       layout = c(Nv, 1),
       par.settings = list(superpose.symbol = list(pch = 1, cex = 0.7))) +
  layer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                      level = 0.5),
        panel.abline(a = 10, b = 1))
```

Using data like these, a researcher may attempt to obtain an estimate of the within-patient effect of dosage by fitting mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models are sequentially nested in their fixed effects: (1) `~ 1`, intercept only; (2) `~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`. The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`. We thus fit four models to four data sets for a total of 16 models.

```{r model-fits,include=FALSE, cache=TRUE}
model.formulas <- c(
  ' ~ 1'             =  y ~ 1 + (1 | patient),
  '~ 1 + x'          =  y ~ 1 + x + (1 | patient),
  '~ 1 + x + xm'     =  y ~ 1 + x + xm + (1 | patient),
  '~ 1 + I(x - xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
)
fits <- lapply(split(Data, ~ SD_ratio),
               function(d) {
                 lapply(model.formulas, function(form) {
                   glmmTMB(form, data = d)
                 })
               })
```

We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---that is, the BLUPs---as would be used for cross-validation based on cases (i.e., occasions within patients):
```{r predict,include=FALSE}
pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  
pred.BLUPs <- lapply(fits, lapply, predict)
```
```{r data-predictions,include=FALSE, cache=TRUE}
Dataf <- lapply(split(Data, ~ SD_ratio),
                function(d) {
                  lapply(names(model.formulas), 
                         function(form) cbind(d, formula = form))
                }) |> 
  lapply(function(dlist) do.call(rbind, dlist)) |> 
  do.call(rbind, args = _) |> 
  within(
    {
      pred.fixed <- unlist(pred.fixed)
      pred.BLUPs <- unlist(pred.BLUPs)
      panel <- factor(formula, levels = c(names(model.formulas), 'data'))
    }
  )
Datap <- reshape(Dataf, direction = "long", sep = '.',
                 varying = c('pred.fixed', 'pred.BLUPs'),
                 timevar = 'prediction_type')
Datap$panel2 <- with(Datap, paste0(formula,":",prediction_type))
flevels <- c(unique(Datap$panel2), 'data')
Datap$panel2 <- factor(Datap$panel2, levels = flevels)
Data$panel <- factor('data', levels = c(names(model.formulas), 'data'))
Data$panel2 <- factor('data', levels = flevels)
```
The fixed- and random-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-combined} along with a summary of the data using 50% concentration ellipses (see the top row of the figure).

@@

{ library("lattice")
  library("latticeExtra")
  xyplot(y ~ x | SD_ratio * panel3, Data,
         groups = patient, type = 'n',
         drop.unused.levels = F,
         par.strip.text = list(cex = 0.4),
         scales = list(x = list(alternating = FALSE, cex = 0.7),
                       y = list(alternating =2, cex = 0.5)),
         xlab = list('dosage', cex = 0.7),
         between =
             list(y = c(0,0,0,0.25,0,0,0,.25)),
         ylab = list(
           paste0(
             paste0(rep(' ',18),  collapse = ''),
                  'fixed-effects',
             paste0(rep(' ',52), collapse = ''),
             'BLUPs',
             paste0(rep(' ',32), collapse = ''),
             'data'),
           cex = 0.7)) +
    glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,
                         level = 0.5),
           panel.abline(a = 10, b = 1)) +
    xyplot(pred  ~ x | SD_ratio * panel2, Datap, type = 'l',
           groups = patient,
           drop.unused.levels = F)
} -> re.plot
useOuterStrips(re.plot)

@@

Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions, and also of BLUPs, towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`. The inclusion of a contextual variable in the model corrects this problem.

Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform to the within-patient slopes for the `~ 1 + x` model in the two data sets with the smallest between-patient variances.

For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variance.  With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.

@@

The third model, `~ 1 + x + xm`, includes a contextual effect of `x`---that is, the cluster mean `xm`---along with `x` and the intercept in the fixed-effect part of the model, and with a random intercept. This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | patient)`, which is the model that generated the data.  The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects-only model with `x` and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).

We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.[^cv-list] The results for all data sets, using both cluster-based and case-based cross-validation, are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.

[^cv-list]: In order to reduce between-model random variability in comparisons of models on the same data set, we applied `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.

```{r echo=FALSE,include=FALSE,cache=TRUE}
library(cv) # unclear why it's necessary to reload cv
.opts <- options(warn = -2) 
```
```{r cross-validation,cache=TRUE,message=FALSE,include=FALSE}
model_lists <- lapply(fits, function(fitlist) do.call(models, fitlist))
cvs_cases <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]], k = 10, 
              data = split(Data, ~ SD_ratio)[[i]])
         })
cvs_clusters <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]],  k = 10, 
              data = split(Data, ~SD_ratio)[[i]], 
              clusterVariables = 'patient')
         })
```

```{r cross-validation-data,include=FALSE,cache=TRUE}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })
dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r cross-validation-data-plot,echo=FALSE}
#| out.width = "75%",
#| fig.height = 3,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(clusters + cases ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', columns = 2, cex = 0.7), 
       type = 'b', ylab = list('CV criterion (MSE)', cex = .7),
       xlab = '',
       par.strip.text = list(cex = 0.7),
       layout= c(Nv, 1),
       par.settings =
         list(superpose.line=list(lty = c(1, 2), lwd = 1),
              superpose.symbol=list(cex = 0.7)),
       scales = list(y = list(log = TRUE, cex = 0.7), 
                     x = list(alternating = F, rot = 45, cex = 0.7))) 
```
In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. With all our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
